---
title: "Dummy coding sample"
author: "Yang Feng"
date: "February 26, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message = FALSE}
# Importing the library
library(dplyr)
```


## Dummy codinng for discrete variables
* **Sample data set**
* **Binary Variable**
* **Variable with three levels**
* **Variable with n levels**
* **Project data set**

### Sample data set
```{r}
# Generating a data set with 100 observations
ID <- c(1 : 100) # Generating IDs
x <- rnorm(100, 0, 20) + 60
y1 <- c(rep("Yes", 20), rep("No", 70), rep(NA, 10))
y2 <- c(rep("Blue", 30), rep("Red", 40), rep("Yellow", 25), rep(NA, 5))
y3 <- c(rep("Extreme Low", 15)， rep("Low", 20), rep("Medium", 30), rep("High", 15), rep("Extreme High", 10), rep(NA, 10))
sample <- data.frame(ID, x, y1, y2, y3)
```

**Glimpse of the sample data set**
```{r echo = FALSE}
glimpse(sample)
```

**First 6 observations**
```{r echo = FALSE}
head(sample)
```

**levels of y1, y2 and y3**
```{r}
summary(sample)
```
As we can see in the summary, there are missing values for y1, y2 and y3

### Binary Variable
```{r}
# Create a dummy variable called y1.d1 for a binary variable
sample$y1.d1 <- ifelse(sample$y1 == "Yes", 1, 0)
```

**Define a function to automate the task**
```{r}
dummy_coding2 <- function(data, response, level, dummy_label){
  dummy <- ifelse(response == level, 1, 0)
  data <- cbind(data, dummy)
  colnames(data)[ncol(data)] <- dummy_label
  data # This function output the dataset after the dummy coding
}
# Let's use this function to automate the task.
data2 <- dummy_coding2(data = sample, response = sample$y1, level = "Yes", dummy_label = "y1.d1_a")
head(data2)
```

**Check y1.d1 and y1.d1_a**
```{r}
# Check y1.d1 and y1.d1_a
neq <- sum(sample$y1.d1!=sample$y1.d1_a)
paste("There are", neq, "different values for y1.d1 and y1.d1_a.")
```


### Variable with three levels
```{r}
# Create two dummy variables called y2.d1 and y2.d2 for y2 (discrete variable with three levels)
# The three levels of y2 are "Blue", "Red" and "Yellow"
# Treating "Red" as the reference level
sample$y2_d1 <- ifelse(sample$y2 == "Blue", 1, 0) # Dummy for Blue
sample$y2_d2 <- ifelse(sample$y2 == "Yellow", 1, 0) # Dummy for Red
# Check
head(sample[,c(4,7,8)])
```

**Define a function to automate the task**
```{r}
dummy_coding3 <- function(data, response, level,  dummy_label){
  col <- colnames(data)
  dummy <- list()
  dummy[[1]] <- ifelse(response == level[1], 1, 0)
  dummy[[2]] <- ifelse(response == level[2], 1, 0)
  data <- cbind(data, dummy[[1]], dummy[[2]])
  colnames(data) <- c(col, dummy_label)
  data
}
# Let's use this function to automate the task.
data3 <- dummy_coding3(data = sample, response = sample$y2, level =c("Blue", "Yellow"), dummy_label = c("y2.d1_a", "y2.d2_a"))
head(data3)
```

**Check y2.d1, y2.d2 and y2.d1_a, y2.d2_a**
```{r echo }
# Check y1.d1 and y1.d1_a
neq2 <- sum(sample$y2.d1!=sample$y2.d1_a & sample$y2_d1 != sample$y2.d2_a)
paste("There are", neq, "different value for y2.d1, y2.d2 and y2.d1_a, y2.d2_a.")
```

### Variable with n levels
```{r}
# Create four dummy variables called y3.d1, y3.d2, y3.d3 and y3.d4 for y3 (discrete variable with five levels)
# The three levels of y2 are "Extreme Low", "Low", "Medium", "High"and "Extreme High"
# Treating "Medium" as the reference level
sample$y3_d1 <- ifelse(sample$y3 == "Extreme Low", 1, 0) # Dummy for Extreme Low
sample$y3_d2 <- ifelse(sample$y3 == "Low", 1, 0) # Dummy for Low
sample$y3_d3 <- ifelse(sample$y3 == "Extreme High", 1, 0) # Dummy for Extreme High
sample$y3_d4 <- ifelse(sample$y3 == "High", 1, 0) # Dummy for High
# Check
head(sample[,c(5,9:12)])
```

**Define a function to automate the task**
```{r}
dummy_coding_n <- function(data, response, level,  dummy_label){
  col_n <- colnames(data)
  dummy <- list()
  for (i in 1:length(level)){
  dummy[[i]] <- ifelse(response == level[i], 1, 0)
  data <- cbind(data, dummy[[i]])
}
  colnames(data) <- c(col_n, dummy_label)
  data
}
# Let's use this function to automate the task.
data3 <- dummy_coding_n(data = sample, response = sample$y3, level =c("Extreme Low", "Low", "Extreme High", "High"), dummy_label = c("y3.d1_a", "y3.d2_a", "y3.d3_a", "y3.d4_a"))
head(data3)
```
Therefore, we can use this function dummy_coding_n to do dummy coding for discrete variables with reference level.

### Project data set
```{r echo = FALSE}
# Importing the dataset
project <- read.csv("D:/Spring 2018/Consulting/Ellita/firenze.csv", header = TRUE)
```

**First 6 observations**
```{r echo = FALSE}
head(project)
```

**Dummy coding for the project data set**
```{r}
# Let's do dummy coding for rs11605924.1., treating GG as the reference group, create dummies rs11605924.1.d1 for TG, rs11605924.1.d2 for TT
project1 <- dummy_coding_n(data = project, response = project$rs11605924.1., level = c("T G", "T T"), dummy_label = c("rs11605924.1.d1", "rs11605924.1.d2"))
head(project1)
```

**Note: When specifying the argument for levels and dummy_label, they are corresponding to each other with respect to order.**

